Research Notes

### 2018-08-27 writeups/notes.tex

>>>>

\chapter{Stratified Flow}

\section{Linear Equations}

We largely follow Booker \& Bretherton from 1966 though do not adopt the
Boussinesq approximation. Consider background shear flow $\vec{u}_0 =
u_0(z)\hat{x}$ and vertical stratification $\rho_0 \propto e^{-z/H}$, then
\begin{subequations}\label{se:shear_eq}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\rho_1}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\rho_1
            - u_{1z}\frac{\rho_0}{H} &= 0,\\
        \pd{u_{1x}}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\p*{u_{1x} + u_0}
            + \frac{1}{\rho_0}\pd{P}{x} &= 0,\\
        \pd{u_{1z}}{t} + \p*{\vec{u} \cdot \vec{\nabla}}u_{1z}
            + \frac{1}{\rho_0}\pd{P}{z} + \frac{\rho_1 g}{\rho_0} &= 0.
    \end{align}
\end{subequations}
Note that $\vec{u} \cdot \vec{\nabla} = \p*{\vec{u}_0 + \vec{u}_1} \cdot
\vec{\nabla}$.

We again assume plane waves in $x, t$ such that $u_{1z}(x, z, t) =
u_{1z}(z)e^{i(k_xx - \omega t)}$. We bracket second-order terms below and set
them to zero in the ensuing calculations
\begin{subequations}
    \begin{align}
        \pd{u_{1z}}{z} + ik_xu_{1x} &= 0,\\
        -i\omega \rho_1 + u_0ik_x\rho_1
            - u_{1z}\frac{\rho_0}{H}
            + \s*{\p*{\vec{u}_1 \cdot \vec{\nabla}} \rho_1} &= 0,\\
        -i\omega u_{1x} + u_0ik_x u_{1x}
            + \frac{1}{\rho_0}ik_xP + u_{1z}\pd{u_0}{z}
            + \s*{\vec{u}_1 \cdot \vec{\nabla} u_{1x}} &= 0,\\
        -i\omega u_{1z} + u_0ik_x u_{1z}
            + \frac{1}{\rho_0}\pd{P}{z} + \frac{\rho_1 g}{\rho_0}
            + \s*{\vec{u}_1 \cdot \vec{\nabla} u_{1z}} &= 0.\\
    \end{align}
\end{subequations}
Denote $\omega_{cr}(z) = \omega - u_0(z)k_x$ the corotating frequency. Using
then that $u_{1x} = \frac{i}{k_x}\pd{u_{1z}}{z}$, $\rho_1 = \frac{i
\rho_0}{H\omega_{cr}} u_{1z}$ and $P = \s*{i\omega_{cr}u_{1x} - u_{1z
\pd{u_0}{z}}} \frac{\rho_0}{ik_x}$, we can finally arrive at linearized equation
\begin{equation}
    \ptd{u_{1z}}{z} - \frac{1}{H}\pd{u_{1z}}{z} + \s*{
        \frac{k_x^2N^2}{\omega_{cr}^2}
            - k_x^2
            - \frac{k_x}{\omega_{cr}H}\pd{u_0}{z}
            + \frac{k_x}{\omega_{cr}} \ptd{u_0}{z}
    } u_{1z} = 0. \label{eq:shear_lin_uz}
\end{equation}
Note that if we make correspondences $ik_x \to \partial_x, i\omega_{cr} \to
\partial_t + u_0\partial_x$ and take $H \to \infty$ while holding $N^2$
constant, we recover Equation 2.1 of Booker \& Bretherton. We can also note that
if we take constant $u_0$ we recover the expected dispersion relation
\autoref{eq:strat_disp} for $\omega = \omega_{cr}$.

\section{Linear Amplification}

One can solve \autoref{eq:shear_lin_uz}

\section{Breaking Criterion}

If the fluid is not strongly stratified $\ptd{u_0}{z}, \pd{u_0}{z}\frac{1}{H}
\ll \frac{k_xN^2}{\omega_{cr}}$ then the singularity is largely induced by
$\omega_{cr} \to 0$ (note that the condition becomes easier to satisfy as we
approach the singularity). $\omega_{cr} = 0$, is exactly the \emph{corotation
resonance}, and one expects that as (note that if $k_z \gg k_x$ then the
dispersion relation is $\omega_{cr} \approx Nk_x/k_z$):
\begin{equation}
    k_z(z) \approx \frac{Nk_x}{\omega_{cr}(z)}
        \sim \frac{\omega_{cr}(z)}{\abs*{u_{1z}}},
        \label{eq:corot_nonlin_cond}
\end{equation}
that nonlinear effects should kick in.

One advantage of a shear flow is that, instead of requiring $2H$ of vertical
height for amplitude growth, we can just choose $u_0 =
\frac{\omega}{k_x}\frac{z}{z_{\max}}$ to meet the corotation citerion at $z =
z_{\max}$. Consider $\pd{u_0}{z} \sim \frac{\omega}{k_xz_{\max}} \approx
\frac{N}{k_zz_{\max}}$. This both satisfies $\pd{u_0}{z} \frac{1}{H} =
\frac{N}{k_zz_{\max}H} \ll \frac{k_xN^2}{\omega_{cr}}$ (as $k_z^2z_{\max}H \ll
1$), and $\ptd{u_0}{z} = 0 \ll \frac{k_xN^2}{\omega_{cr}}$. We can even induce
the corotation resonance well below $z_{\max}$ and still be perfectly within the
regime of validity of our assumption.

A second advantage of a shear flow is that we can force at very small amplitudes
and still observe resonances, since the resonance is driven by the convergence
of $\omega_{cr} \to 0$ (purely coordinate driven) instead of $u_{1z} \sim
v_{ph}$ (dependent on wave amplitude).

The width of the dissipation zone can be estimated by noticing $u_0k_x = \omega
\pm \sqrt{k_xNu_{1z}}$ from the nonlinearity criterion
\autoref{eq:corot_nonlin_cond}. Thus, the width of $u_0$ that incurs nonlinear
effects is roughly of order $2\sqrt{Nu_{1z} / k_x}$, which has the right scaling
behavior (a stronger perturbation should be resonant farther away).
